V4 Neural Network Model for Shape-based Feature Extraction and Object Discrimination

### 2015-09-06 article.tex
finish revision

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{images/fig1.jpg}} 
\caption{V4 is an intermediate stage in visual recognition.}
\label{fig:1}
\end{figure}

@@

\begin{figure}
\centering
\subfloat[]{\includegraphics[width=0.18\linewidth]{images/fig2a.jpg}\label{fig:2a}}\hfil
\subfloat[]{\includegraphics[width=0.36\linewidth]{images/fig2b.jpg}\label{fig:2b}}\hfil
\subfloat[]{\includegraphics[width=0.36\linewidth]{images/fig2c.jpg}\label{fig:2c}}
\caption{Shapes to examine V4 selectivity.
(a) Sinusoidal gratings. (b) Non-Cartesian gratings. (c) Closed shapes.}
\label{fig:2}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=\linewidth]{images/fig3.eps}} 
\caption{Multilayer neural network model of V4.}
\label{fig:3}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{images/fig4.jpg}} 
\caption{Entropy of image patches.
Bar charts in the second row show the output values of complex cells
with different preferred orientations (from $0^\circ$ to $180^\circ$).}
\label{fig:4}
\end{figure}

@@

\begin{figure}
\centering
\subfloat[Car image]{\includegraphics[width=0.32\linewidth]{images/fig5a.jpg}\label{fig:5a}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig5c.jpg}\label{fig:5b}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig5e.jpg}\label{fig:5c}}\\
\subfloat[Face image]{\includegraphics[width=0.32\linewidth]{images/fig5b.jpg}\label{fig:5d}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig5d.jpg}\label{fig:5e}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig5f.jpg}\label{fig:5f}}\\
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig5g.jpg}\label{fig:5g}}\hfil
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig5h.jpg}\label{fig:5h}}
\caption{Feature detection. (a)(d) Original images.
(b)(e) Complex cell activity map. 
(c)(f) Entropy map.
In the maps, red color indicates high value and blue color indicates low value.
(g)(h) Salient points at the scale of 4, 8, and 16 in terms of simple cell radius.}
\label{fig:5}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.99\linewidth]{images/fig6.eps}} 
\caption{RBM encoder for local shape feature.}
\label{fig:6}
\end{figure}

@@

\begin{figure}
\centering
\subfloat[Response map]{\includegraphics[width=0.48\linewidth]{images/fig7a.jpg}\label{fig:7a}}\hfil
\subfloat[Weight matrices]{\includegraphics[width=0.48\linewidth]{images/fig7b.jpg}\label{fig:7b}}
\caption{Response map and weight matrices of a perceptron that distinguishes two shapes.
(a) Responses of the perceptron over 18 samples.
The perceptron prefers the shape in the left half of the samples.
It is insensitive to stimulus position.
(b) The input weight of the perceptron. 
Each block shows the weight of connections from complex cells of a certain orientation.
Complex cells of 18 different orientations provide input for this perceptron.}
\label{fig:7}
\end{figure}

@@

\begin{table}[h]
\caption{Shape selectivity of our model}
\centering
\begin{tabular}{rrr}
\toprule
Selectivity & \#Units & Percentage \\
\midrule
Sinusoidal gratings & 50 & 19.5\% \\
Non-Cartesian gratings & 55 & 21.5\% \\
Segmented curves & 102 & 39.8\% \\
Closed shapes & 68 & 26.6\% \\
\bottomrule
\end{tabular}
\label{tab:1}
\end{table}

@@

\begin{figure}
\centering
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8a.jpg}\label{fig:8a}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8b.jpg}\label{fig:8b}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8c.jpg}\label{fig:8c}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8d.jpg}\label{fig:8d}}
\caption{RBM output units responding to convex curves towards certain directions.
(a) Sample patterns of segmented curves.
(b-d) Selectivity indexes and response maps of units tuned for upward, downward, and top-left projections. }
\label{fig:8}
\end{figure}

\begin{figure}
\centering
\subfloat[Response map of a unit selective for a pair of circular sectors (selectivity index = 3.92).]%
{\includegraphics[width=0.9\linewidth]{images/fig9a.jpg}\label{fig:9a}}\\
\subfloat[Response map of a unit selective for helix shapes (Selectivity index = 6.37).]%
{\includegraphics[width=0.9\linewidth]{images/fig9b.jpg}\label{fig:9b}}
\caption{Response maps of RBM units tuned for non-Cartesian gratings.}
\label{fig:9}
\end{figure}

\begin{figure}
\centering
\subfloat[Patterns]{\includegraphics[width=0.31\linewidth]{images/fig10a.jpg}\label{fig:10a}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig10b.jpg}\label{fig:10b}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig10c.jpg}\label{fig:10c}}
\caption{RBM units responding to closed shapes.}
\label{fig:10}
\end{figure}

@@

We first provide a qualitative evaluation.
The SIFT features are used for comparison.
The VLFeat open-source library \cite{vedaldi2010} were utilized for the SIFT implementation and the feature matching algorithm.

\begin{figure}
\centering
  \subfloat[Matching between the same face]{
    \includegraphics[width=0.8\linewidth]{images/fig11a.jpg}
    \label{fig:11a}}\\
  \subfloat[Matching between different faces]{
    \includegraphics[width=0.8\linewidth]{images/fig11b.jpg}
    \label{fig:11b}}
\caption{Matching face images.}
\label{fig:11}
\end{figure}

@@

\begin{figure}
\centering
  \subfloat[Our match]{
    \includegraphics[width=0.8\linewidth]{images/fig12a.jpg}
    \label{fig:12a}}\\
  \subfloat[SIFT match]{
    \includegraphics[width=0.8\linewidth]{images/fig12b.jpg}
    \label{fig:12b}}
\caption{Comparison of feature matching between van images.}
\label{fig:12}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{images/fig13.jpg}} 
\caption{Matching facial features in Caltech-101 ``Face-easy'' category.}
\label{fig:13}
\end{figure}

@@

\begin{figure}
\centering
  \subfloat[Sampling density of orientations]{
    \includegraphics[width=0.9\linewidth]{images/fig14a.eps}
    \label{fig:14a}}\\
  \subfloat[Number of features per image]{
    \includegraphics[width=0.9\linewidth]{images/fig14b.eps}
    \label{fig:14b}}
\caption{Performance with different parameters.}
\label{fig:14}
\end{figure}

@@

\begin{figure}
\centering
  \subfloat[Comparison with other methods]{
    \includegraphics[width=0.9\linewidth]{images/fig15a.eps}
    \label{fig:15a}}\\
  \subfloat[Comparison with other features]{
    \includegraphics[width=0.9\linewidth]{images/fig15b.eps}
    \label{fig:15b}}
\caption{Comparison of performance.}
\label{fig:15}
\end{figure}

@@

\subsection{Traditional image features}

The model proposed in this paper is a biologically motivated model for image feature extraction.
As reviewed in \cite{mikolajczyk2005}, traditional image features fall into three categories: 
distribution-based descriptors, spatial-frequency descriptors and differential-based descriptors.
Distribution-based descriptors use histograms to represent characteristics of local image regions.
SIFT descriptor \cite{lowe1999} and shape context \cite{belongie2002} are typical examples of such descriptors.
Spatial-frequency descriptors and differential-based descriptors use filters and derivatives to capture the invariant features. 
Examples of these categories are complex filters \cite{schaffalitzky2002} and Baumberg's affine invariant descriptor \cite{baumberg2000}.
These traditional image features utilize flat models which compute features directly from the image via statistical or algebraic techniques. 
They have little in common with the neural systems.
On the contrary, our proposed model simulates the biological vision in the level of neural mechanism.
It is a multilayer model which employs the hierarchical structure of neural systems and extracts information from simple features to gradually complex features.

This process requires a large number of training samples (\cite{krizhevsky2012} uses 1.2 million training images)
and a neural network of very large scale (In \cite{krizhevsky2012}, the network has 8 layers and more than 400,000 neurons).
Deep learning can be viewed as automatic feature extraction.
In our proposed model, we combine automatic feature extraction with artificial feature extraction.
With prior knowledge of the cortical neurons, we manually design the layers corresponding to the primary visual cortex
and V4 features are automatically extracted based on the output of the artificially designed network.
This strategy saves us from retraining the first layers every time
and it allows the model to work on limited number of training images.

In this paper, we propose a novel model for the visual area V4.
It is based on the neural mechanism of the ventral visual pathway.
We focus on visual attention and shape selectivity in V4.
V4 in our model is implemented as a multilayer neural network 
which selects salient points and encodes the shape feature in the neighborhood of the salient points.
We demonstrate a variety of experiments in which the model was evaluated.
The results show that the proposed model is consistent with the shape selectivity of area V4.
It can find salient points in images and encode local shape feature into a discriminative representation.

We further evaluate our model in object recognition experiment.
We designed a image descriptor which synthesizes several V4 computation units
into an IT unit to produce a set of invariant local image features.
Compared with traditional image descriptors,
the proposed model shows competitive performance in object recognition experiment. 
It captures the essential shape feature of objects
and thus achieves a good balance between discrimination and generalization.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{images/fig-17.jpg}
\caption{Framework of the proposed model.}
\label{fig:17}
\end{figure}

The framework of the proposed multilayer neural network model is shown in \figurename~\ref{fig:17}.
Our contribution is focused on the V4 layer,
where salient shape features are detected and located,
and the features are encoded and further synthesized to produce discriminative representation of object shapes.
The proposed model demonstrates a successful attempt to construct model of the ventral visual pathway
in the brain for visual discrimination.
We explicitly adopt the model of several layers of this pathway,
including well studied lower-level layers, such as simple cells and complex cells.
We focus our work on V4, which is significant but lacks mature computation models.

Future work should involve improvement of the IT unit and further evaluation on learning and detection shapes.
The implementation of the IT unit should conform to neural mechanism
and should be capable of learning generalized representation of shape categories.
Information about color and texture of objects is also very important,
which should also participate in the process of object discrimination.
These show the direction of our future work.

>>>>

\begin{figure}[!t]
\centerline{\includegraphics[width=0.8\linewidth]{images/fig1.jpg}} 
\caption{V4 is an intermediate stage in visual recognition.}
\label{fig:1}
\end{figure}

@@

\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.18\linewidth]{images/fig2a.jpg}\label{fig:2a}}\hfil
\subfloat[]{\includegraphics[width=0.36\linewidth]{images/fig2b.jpg}\label{fig:2b}}\hfil
\subfloat[]{\includegraphics[width=0.36\linewidth]{images/fig2c.jpg}\label{fig:2c}}
\caption{Shapes to examine V4 selectivity.
(a) Sinusoidal gratings. (b) Non-Cartesian gratings. (c) Closed shapes.}
\label{fig:2}
\end{figure}

@@

\begin{figure}[!t]
\centerline{\includegraphics[width=\linewidth]{images/fig3.eps}} 
\caption{Multilayer neural network model of V4.}
\label{fig:3}
\end{figure}

@@

\begin{figure}[!t]
\centerline{\includegraphics[width=0.8\linewidth]{images/fig4.jpg}} 
\caption{Entropy of image patches.
Bar charts in the second row show the output values of complex cells
with different preferred orientations (from $0^\circ$ to $180^\circ$).}
\label{fig:4}
\end{figure}

@@

\begin{figure}[!t]
\centering
\subfloat[Car image]{\includegraphics[width=0.32\linewidth]{images/fig5a.jpg}\label{fig:5a}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig5c.jpg}\label{fig:5b}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig5e.jpg}\label{fig:5c}}\\
\subfloat[Face image]{\includegraphics[width=0.32\linewidth]{images/fig5b.jpg}\label{fig:5d}}\hfil
\subfloat[Activity map]{\includegraphics[width=0.32\linewidth]{images/fig5d.jpg}\label{fig:5e}}\hfil
\subfloat[Entropy map]{\includegraphics[width=0.32\linewidth]{images/fig5f.jpg}\label{fig:5f}}\\
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig5g.jpg}\label{fig:5g}}\hfil
\subfloat[Salient points]{\includegraphics[width=0.49\linewidth]{images/fig5h.jpg}\label{fig:5h}}
\caption{Feature detection. (a)(d) Original images.
(b)(e) Complex cell activity map. 
(c)(f) Entropy map.
In the maps, red color indicates high value and blue color indicates low value.
(g)(h) Salient points at the scale of 4, 8, and 16 in terms of simple cell radius.}
\label{fig:5}
\end{figure}

@@

\begin{figure}[!t]
\centerline{\includegraphics[width=0.99\linewidth]{images/fig6.eps}} 
\caption{RBM encoder for local shape feature.}
\label{fig:6}
\end{figure}

@@

\begin{figure}[!t]
\centering
\subfloat[Response map]{\includegraphics[width=0.48\linewidth]{images/fig7a.jpg}\label{fig:7a}}\hfil
\subfloat[Weight matrices]{\includegraphics[width=0.48\linewidth]{images/fig7b.jpg}\label{fig:7b}}
\caption{Response map and weight matrices of a perceptron that distinguishes two shapes.
(a) Responses of the perceptron over 18 samples.
The perceptron prefers the shape in the left half of the samples.
It is insensitive to stimulus position.
(b) The input weight of the perceptron. 
Each block shows the weight of connections from complex cells of a certain orientation.
Complex cells of 18 different orientations provide input for this perceptron.}
\label{fig:7}
\end{figure}

@@

\begin{table}[h]
\caption{Shape selectivity of our model}
\centering
\begin{tabular}{|r|r|r|}
\hline
Selectivity & \#Units & Percentage \\\hline
Sinusoidal gratings & 50 & 19.5\% \\\hline
Non-Cartesian gratings & 55 & 21.5\% \\\hline
Segmented curves & 102 & 39.8\% \\\hline
Closed shapes & 68 & 26.6\% \\\hline
\end{tabular}
\label{tab:1}
\end{table}

@@

\begin{figure}[!t]
\centering
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8a.jpg}\label{fig:8a}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8b.jpg}\label{fig:8b}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8c.jpg}\label{fig:8c}}\hfil
\subfloat[]{\includegraphics[width=0.24\linewidth]{images/fig8d.jpg}\label{fig:8d}}
\caption{RBM output units responding to convex curves towards certain directions.
(a) Sample patterns of segmented curves.
(b-d) Selectivity indexes and response maps of units tuned for upward, downward, and top-left projections. }
\label{fig:8}
\end{figure}

\begin{figure}[!t]
\centering
\subfloat[Response map of a unit selective for a pair of circular sectors (selectivity index = 3.92).]%
{\includegraphics[width=0.9\linewidth]{images/fig9a.jpg}\label{fig:9a}}\\
\subfloat[Response map of a unit selective for helix shapes (Selectivity index = 6.37).]%
{\includegraphics[width=0.9\linewidth]{images/fig9b.jpg}\label{fig:9b}}
\caption{Response maps of RBM units tuned for non-Cartesian gratings.}
\label{fig:9}
\end{figure}

\begin{figure}[!t]
\centering
\subfloat[Patterns]{\includegraphics[width=0.31\linewidth]{images/fig10a.jpg}\label{fig:10a}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig10b.jpg}\label{fig:10b}}\hfil
\subfloat[Response map]{\includegraphics[width=0.31\linewidth]{images/fig10c.jpg}\label{fig:10c}}
\caption{RBM units responding to closed shapes.}
\label{fig:10}
\end{figure}

@@

We first provide a qualitative evaluation.
The SIFT features \cite{lowe1999} are used for comparison.
The VLFeat open-source library \cite{vedaldi2010} were utilized for the SIFT implementation and the feature matching algorithm.

\begin{figure}[!t]
\centering
  \subfloat[Matching between the same face]{
    \includegraphics[width=0.8\linewidth]{images/fig11a.jpg}
    \label{fig:11a}}\\
  \subfloat[Matching between different faces]{
    \includegraphics[width=0.8\linewidth]{images/fig11b.jpg}
    \label{fig:11b}}
\caption{Matching face images.}
\label{fig:11}
\end{figure}

@@

\begin{figure}[!t]
\centering
  \subfloat[Our match]{
    \includegraphics[width=0.8\linewidth]{images/fig12a.jpg}
    \label{fig:12a}}\\
  \subfloat[SIFT match]{
    \includegraphics[width=0.8\linewidth]{images/fig12b.jpg}
    \label{fig:12b}}
\caption{Comparison of feature matching between van images.}
\label{fig:12}
\end{figure}

@@

\begin{figure}[!t]
\centerline{\includegraphics[width=0.9\linewidth]{images/fig13.eps}} 
\caption{Matching facial features in Caltech-101 ``Face-easy'' category.}
\label{fig:13}
\end{figure}

@@

\begin{figure}[!t]
\centering
  \subfloat[Sampling density of orientations]{
    \includegraphics[width=0.9\linewidth]{images/fig14a.eps}
    \label{fig:14a}}\\
  \subfloat[Number of features per image]{
    \includegraphics[width=0.9\linewidth]{images/fig14b.eps}
    \label{fig:14b}}
\caption{Performance with different parameters.}
\label{fig:14}
\end{figure}

@@

\begin{figure}[!t]
\centering
  \subfloat[Comparison with other methods]{
    \includegraphics[width=0.9\linewidth]{images/fig15a.eps}
    \label{fig:15a}}\\
  \subfloat[Comparison with other features]{
    \includegraphics[width=0.9\linewidth]{images/fig15b.eps}
    \label{fig:15b}}
\caption{Comparison of performance.}
\label{fig:15}
\end{figure}

@@

\begin{table*}[!t]
\caption{Comparison of different models}
\centering
\small
\begin{tabular}{|p{0.15\linewidth}|p{0.22\linewidth}|p{0.22\linewidth}|p{0.22\linewidth}|}
\hline
& Our model & HMAX & Deep learning \\\hline
Number of layers & 3 & 2 & $2\sim8$ \\
& ($+$ saliency filters) & ($+2$ pooling layers) & ($+2\sim3$ pooling layers)\\\hline
Input images & various size & various size & fixed size \\
& (multi-scale salient point detector) & (full-scan with fixed-size window) & \\\hline
Training samples & $5\sim50$ & $5\sim50$ & $>10^4$ \\
& (per category) & (per category) & (for unsupervised pre-training) \\\hline
Training & restricted Boltzmann & greedy search for fitting & restricted Boltzmann \\
algorithms & machine (only V4 layer) & V4 neuron; radial basis function network for image feature extraction & machine (all layers) \\\hline
\end{tabular}
\label{tab:2}
\end{table*}

The model proposed in this paper is a biologically motivated model for shape-based image feature extraction.
Traditional image descriptors utilize flat models which extract features directly from the image via statistical or algebraic techniques \cite{mikolajczyk2005}. 
On the contrary, our proposed model simulates the biological vision in the level of neural mechanism.
Its multilayer structure resembles the neural systems and extracts information from simple features to gradually more complex features.

Previous work on biologically inspired models shows that efficient object recognition algorithms can be developed 
based on the representation generated by the primary visual cortex \cite{wei2014,wei2014b,zhao2013}.
In this paper, we investigate deeper into higher levels of the visual cortex.
We simulate the shape selectivity in area V4 because the shape or contour of an object is generally stable and invariant, which provides a good basis for recognition.

The HMAX model \cite{cadieu2007} and the deep learning model \cite{bengio2009}, which both draw inspiration from the neural systems, 
construct similar multilayer structures.
In particular, the deep model exhibits high performance in classifying very large image database \cite{krizhevsky2012}.
However, a deep model is a neural network of very large scale (In \cite{krizhevsky2012}, the network has 8 layers and more than 400,000 neurons),
and training such a model generally requires a large number of training samples (\cite{krizhevsky2012} uses 1.2 million training images).
Deep learning is a process of fully automatic feature extraction.
Limited training samples can not provide enough variety to keep the model from over-fitting.
The HMAX model consists of fewer layers and only one layer (S2 layer) is trainable, which allows the model to work on limited train samples.
In our proposed model, we combine these two approach.
With prior knowledge of the cortical neurons, we manually design the layers corresponding to the primary visual cortex
and V4 features are automatically extracted based on the output of the artificially designed network.
This strategy saves us from retraining the first layers every time.
Additionally, we implement a layer of saliency filters to detect key-points in images,
which enables our model to process images of various sizes efficiently. 
In \tablename~\ref{tab:2}, we summarize the comparison of the three models.

Future work should involve further investigation into the visual systems and extensive evaluation on shape-based object detection.
Currently the model extracts only shape-based features.
Information about color and texture is also very important, which should also participate in the process of visual recognition.
The proposed model is a feed-forward network.
We should also consider the function of feedback connections in the neural systems.
These show the direction of our future work.
